Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold Standard Items

### 2025-06-30 analysis/primary_analyses.qmd
update tbl-2, tbl-5 and description in results; format notebook html

---
title: "Primary analyses"
---

@@

fit_df <- bind_rows(
  data.frame(as.list(fitMeasures(fit1A6P))) |> mutate(measure = "age", categories = "6"),
  data.frame(as.list(fitMeasures(fit1A2P))) |> mutate(measure = "age", categories = "2"),
  data.frame(as.list(fitMeasures(fit1G6P))) |> mutate(measure = "height", categories = "6"),
  data.frame(as.list(fitMeasures(fit1G2P))) |> mutate(measure = "height", categories = "2"),
  data.frame(as.list(fitMeasures(fit1Ge6P))) |> mutate(measure = "weight", categories = "6"),
  data.frame(as.list(fitMeasures(fit1Ge2P))) |> mutate(measure = "weight", categories = "2")
) |>
  pivot_longer(!c(measure, categories), names_to = "statistic", values_to = "est") |>
  filter(statistic %in% c("chisq.scaled", "pvalue.scaled", "df.scaled", "cfi.scaled", "rmsea.scaled", "srmr")) |>
  mutate(statistic = case_match(statistic, "chisq.scaled" ~ "$\\chi^2$", "pvalue.scaled" ~ "p", "df.scaled" ~ "df", "cfi.scaled" ~ "CFI", "rmsea.scaled" ~ "RMSEA", "srmr" ~ "SRMR")) |>
  mutate(est_print = papaja::printnum(est, gt1 = FALSE, digits = 3)) |>
  pivot_wider(id_cols = statistic, names_from = c(measure, categories), values_from = est_print)

@@

  flextable::bold(i = ~ age_6 > age_6[14], j = "age_6") |>
  flextable::bold(i = ~ age_2 > age_2[14], j = "age_2") |>
  flextable::bold(i = ~ height_6 > height_6[14], j = "height_6") |>
  flextable::bold(i = ~ height_2 > height_2[14], j = "height_2") |>
  flextable::bold(i = ~ weight_6 > weight_6[14], j = "weight_6") |>
  flextable::bold(i = ~ weight_2 > weight_2[14], j = "weight_2") |>
  flextable::bold(i = 15:21, bold = FALSE) |>
  flextable::bg(i  = 14, bg = "grey90") |>

>>>>

---
title: "Primary analyses"
format:
  html:
    theme:
      - cosmo
      - ../notebooks.scss
    grid:
      body-width: 900px
---

@@

fit_df <- bind_rows(
  data.frame(as.list(fitMeasures(fit1A6P))) |> mutate(measure = "age", categories = "6"),
  data.frame(as.list(fitMeasures(fit1A2P))) |> mutate(measure = "age", categories = "2"),
  data.frame(as.list(fitMeasures(fit1G6P))) |> mutate(measure = "height", categories = "6"),
  data.frame(as.list(fitMeasures(fit1G2P))) |> mutate(measure = "height", categories = "2"),
  data.frame(as.list(fitMeasures(fit1Ge6P))) |> mutate(measure = "weight", categories = "6"),
  data.frame(as.list(fitMeasures(fit1Ge2P))) |> mutate(measure = "weight", categories = "2")
) |>
  pivot_longer(!c(measure, categories), names_to = "statistic", values_to = "est") |>
  filter(statistic %in% c("chisq.scaled", "pvalue.scaled", "df.scaled", "cfi.scaled", "rmsea.scaled", "srmr")) |>
  mutate(statistic = case_match(statistic, "chisq.scaled" ~ "$\\chi^2$", "pvalue.scaled" ~ "p", "df.scaled" ~ "df", "cfi.scaled" ~ "CFI", "rmsea.scaled" ~ "RMSEA", "srmr" ~ "SRMR")) |>
  mutate(est_print = papaja::printnum(est, gt1 = FALSE, digits = 3)) |>
  mutate(est_print = if_else(statistic == "df", as.character(est), est_print)) |>
  mutate(est_print = if_else(statistic == "p", papaja::printp(as.numeric(est_print)), est_print)) |>
  pivot_wider(id_cols = statistic, names_from = c(measure, categories), values_from = est_print)

@@

  flextable::bold(i = ~ abs(parse_number(age_6)) > parse_number(age_6[14]), j = "age_6") |>
  flextable::bold(i = ~ abs(parse_number(age_2)) > parse_number(age_2[14]), j = "age_2") |>
  flextable::bold(i = ~ abs(parse_number(height_6)) > parse_number(height_6[14]), j = "height_6") |>
  flextable::bold(i = ~ abs(parse_number(height_2)) > parse_number(height_2[14]), j = "height_2") |>
  flextable::bold(i = ~ abs(parse_number(weight_6)) > parse_number(weight_6[14]), j = "weight_6") |>
  flextable::bold(i = ~ abs(parse_number(weight_2)) > parse_number(weight_2[14]), j = "weight_2") |>
  flextable::bold(i = 15:21, bold = FALSE) |>
  flextable::bg(i  = 14, bg = "grey90") |>

### 2025-06-30 analysis/secondary_analyses.qmd
update tbl-2, tbl-5 and description in results; format notebook html

---
title: "Secondary analyses"
---

@@

fit_df <- bind_rows(
  data.frame(as.list(fitMeasures(fit1A6))) |> mutate(measure = "age", categories = "6"),
  data.frame(as.list(fitMeasures(fit1A2))) |> mutate(measure = "age", categories = "2"),
  data.frame(as.list(fitMeasures(fit1G6))) |> mutate(measure = "height", categories = "6"),
  data.frame(as.list(fitMeasures(fit1G2))) |> mutate(measure = "height", categories = "2"),
  data.frame(as.list(fitMeasures(fit1Ge6))) |> mutate(measure = "weight", categories = "6"),
  data.frame(as.list(fitMeasures(fit1Ge2))) |> mutate(measure = "weight", categories = "2")
) |>
  pivot_longer(!c(measure, categories), names_to = "statistic", values_to = "est") |>
  filter(statistic %in% c("chisq.scaled", "pvalue.scaled", "df.scaled", "cfi.scaled", "rmsea.scaled", "srmr")) |>
  mutate(statistic = case_match(statistic, "chisq.scaled" ~ "$\\chi^2$", "pvalue.scaled" ~ "p", "df.scaled" ~ "df", "cfi.scaled" ~ "CFI", "rmsea.scaled" ~ "RMSEA", "srmr" ~ "SRMR")) |>
  mutate(est_print = papaja::printnum(est, gt1 = FALSE, digits = 3)) |>
  pivot_wider(id_cols = statistic, names_from = c(measure, categories), values_from = est_print)

>>>>

---
title: "Secondary analyses"
format:
  html:
    theme:
      - cosmo
      - ../notebooks.scss
    grid:
      body-width: 900px
---

@@

fit_df <- bind_rows(
  data.frame(as.list(fitMeasures(fit1A6))) |> mutate(measure = "age", categories = "6"),
  data.frame(as.list(fitMeasures(fit1A2))) |> mutate(measure = "age", categories = "2"),
  data.frame(as.list(fitMeasures(fit1G6))) |> mutate(measure = "height", categories = "6"),
  data.frame(as.list(fitMeasures(fit1G2))) |> mutate(measure = "height", categories = "2"),
  data.frame(as.list(fitMeasures(fit1Ge6))) |> mutate(measure = "weight", categories = "6"),
  data.frame(as.list(fitMeasures(fit1Ge2))) |> mutate(measure = "weight", categories = "2")
) |>
  pivot_longer(!c(measure, categories), names_to = "statistic", values_to = "est") |>
  filter(statistic %in% c("chisq.scaled", "pvalue.scaled", "df.scaled", "cfi.scaled", "rmsea.scaled", "srmr")) |>
  mutate(statistic = case_match(statistic, "chisq.scaled" ~ "$\\chi^2$", "pvalue.scaled" ~ "p", "df.scaled" ~ "df", "cfi.scaled" ~ "CFI", "rmsea.scaled" ~ "RMSEA", "srmr" ~ "SRMR")) |>
  mutate(est_print = papaja::printnum(est, gt1 = FALSE, digits = 3)) |>
  mutate(est_print = if_else(statistic == "df", as.character(est), est_print)) |>
  mutate(est_print = if_else(statistic == "p", papaja::printp(as.numeric(est_print)), est_print)) |>
  pivot_wider(id_cols = statistic, names_from = c(measure, categories), values_from = est_print)

### 2025-06-30 manuscript.qmd
mention EFA analyses in ESM
update tbl-2, tbl-5 and description in results; format notebook html

First, we constructed items for the physical characteristics of body height, body weight, and age. Age was defined as chronological age or age in years, corresponding to the time elapsed since a person's birth [@montepareYouReOnly1989; @schwallDefiningAgeUsing2012]. Height was defined as the height of an upright person from the sole of the foot to the top of the head in centimeters, and body weight as the physical mass of a person in kilograms [@martinAnthropometrieAnleitungSelbstaendigen1929]. The items were developed by 124 psychology students [deductive item construction, @burischApproachesPersonalityInventory1984] and 24 persons [prototype approach, see @broughtonPrototypeStrategyConstruction1984] who deviate from the German population mean at least one standard deviation in the relevant characteristics (separately for women and men). The 24 persons were asked to think of prototypical behaviors for each construct to assess the whole spectrum of the latent variable. This process produced a total of 138 items (response categories were not tested). In a second step, we examined how these items were interpreted and whether they were connected to ideas unrelated to the constructs using cognitive interviews. We conducted eight interviews with a length of two to three hours and applied various cognitive procedures, including probing, paraphrasing, concurrent-think-aloud, and retrospective-think-aloud [@prueferVerfahrenEvaluationSurvey1996; @prueferKognitiveInterviews2005]. This should be a sufficient number as recommendations suggest five to 30 interviews and indicate that the most serious problems can already be identified with a small number of interviews [@willisCognitiveInterviewing2005]. The interviews were conducted with `r nrow(dat_interviews)` people from the target group (`r nrow(dat_interviews[dat_interviews$Geschlecht == 1,])` women and `r nrow(dat_interviews[dat_interviews$Geschlecht == 2,])` men) aged between `r min(dat_interviews$Alter)` and `r max(dat_interviews$Alter)` (*M* = `r printnum(mean(dat_interviews$Alter))`, *SD* = `r printnum(sd(dat_interviews$Alter))`). The level of education ranged from high school diploma to university degree. After the cognitive interviews, misleading items were reformulated or eliminated [@faulbaumWasIstGute2009], resulting in 61 remaining items. In a third step, these 61 items were tested online on a new sample, which initially consisted of 456 people. However, we excluded `r 456 - nrow(dat_pilot)` subjects because they had not completed the questionnaire, resulting in a sample of `r nrow(dat_pilot)` participants aged between `r min(dat_pilot$Alter)` and `r max(dat_pilot$Alter)` years (*M* = `r printnum(mean(dat_pilot$Alter))`, *SD* = `r printnum(sd(dat_pilot$Alter))`), including `r nrow(dat_pilot[dat_pilot$Geschlecht == 1,])` women (`r printnum(nrow(dat_pilot[dat_pilot$Geschlecht == 1,]) / nrow(dat_pilot) * 100)` %) and `r nrow(dat_pilot[dat_pilot$Geschlecht == 2,])` men (`r printnum(nrow(dat_pilot[dat_pilot$Geschlecht == 2,]) / nrow(dat_pilot) * 100)` %). The level of education was distributed as follows: `r round(nrow(dat_pilot[dat_pilot$Bildungsniveau == 1,]) / nrow(dat_pilot) * 100, 2)` % no school leaving certificate, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 2,]) / nrow(dat_pilot) * 100)` % secondary school leaving certificate/elementary school or equivalent, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 3,]) / nrow(dat_pilot) * 100)` % secondary school or equivalent, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 4,]) / nrow(dat_pilot) * 100)` % vocational baccalaureate or high school diploma, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 5,]) / nrow(dat_pilot) * 100)` % college degree or university degree, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 6,]) / nrow(dat_pilot) * 100)` % doctorate or habilitation. Out of all respondents, `r printnum(nrow(dat_pilot[dat_pilot$Muttersprache == 1,]) / nrow(dat_pilot) * 100)` % stated German as their native language. Based on these survey responses, we excluded items if a) their main item loading was not on the intended factor, b) their main and secondary loadings were almost equal, or c) they had loading below or equal to .30. Additionally, if (d) items had very similar content, items with lower loadings were excluded. Based on these criteria 12 items were selected for each construct (see @tbl-5). For each of the three questionnaires, two versions were created – one with a 6-point scale with verbal endpoints ("does not apply" to "applies", only endpoints were labeled) and fully numerical anchoring for the dichotomous response format with "applies" or "does not apply".

@@

Both scale lengths exhibited similar loading patterns, but those with two response categories obtained slightly higher loadings compared to the six response categories in most cases. On average, questionnaires with two response categories showed better descriptive fit indices than those with six. Model fits were within the expected range for questionnaire items [@goretzkoEvaluatingModelFit2024]. In particular, the fit indices for the questionnaires on body height and weight with two response categories showed the best fit.

```{r}
#| eval: false

# height
gold_loading_h6 <- results$load_df_nomod |>
  filter(categories == 6, measure == "height", rhs == "Koerpergroesse_realW") |>
  pull(est.std)
higher_load_h6 <- results$load_df_nomod |> 
  filter(categories == 6, str_detect(results$load_df_nomod$item_name, "KGr_")) |>
  summarize(higher_load = sum(est.std > gold_loading_h6))

gold_loading_h2 <- results$load_df_nomod |>
  filter(categories == 2, measure == "height", rhs == "Koerpergroesse_realW") |>
  pull(est.std)
higher_load_h2 <- results$load_df_nomod |> 
  filter(categories == 2, str_detect(results$load_df_nomod$item_name, "KGr_")) |>
  summarize(higher_load = sum(est.std > gold_loading_h2))

# weight
gold_loading_w6 <- results$load_df_nomod |>
  filter(categories == 6, measure == "weight", rhs == "Koerpergewicht_realW") |>
  pull(est.std)
higher_load_w6 <- results$load_df_nomod |> 
  filter(categories == 6, str_detect(results$load_df_nomod$item_name, "KGe_")) |>
  summarize(higher_load = sum(est.std > gold_loading_w6))

gold_loading_w2 <- results$load_df_nomod |>
  filter(categories == 2, measure == "weight", rhs == "Koerpergewicht_realW") |>
  pull(est.std)
higher_load_w2 <- results$load_df_nomod |> 
  filter(categories == 2, str_detect(results$load_df_nomod$item_name, "KGe_")) |>
  summarize(higher_load = sum(est.std > gold_loading_w2))

# age
gold_loading_a6 <- results$load_df_nomod |>
  filter(categories == 6, measure == "weight", rhs == "Alter_realW") |>
  pull(est.std)
higher_load_a6 <- results$load_df_nomod |> 
  filter(categories == 6, str_detect(results$load_df_nomod$item_name, "A_")) |>
  summarize(higher_load = sum(est.std > gold_loading_a6))

gold_loading_a2 <- results$load_df_nomod |>
  filter(categories == 2, measure == "age", rhs == "Alter_realW") |>
  pull(est.std)
higher_load_a2 <- results$load_df_nomod |> 
  filter(categories == 2, str_detect(results$load_df_nomod$item_name, "A_")) |>
  summarize(higher_load = sum(est.std > gold_loading_a2))
```

```{r}
higher_load <- results$load_df_nomod |> group_by(measure, categories) |>
  reframe(freq = mean(head(abs(est.std), 12) > tail(abs(est.std), 1)))
```

For the body height scales, the self-reported physical height loaded highest (six response categories) or similarly high (two response categories) on the latent variables compared to the remaining items. However, that was not the case for the weight and age scales.

For weight, the items 7 ("My pants have a small waistband.") and 4 ("I am obese.") had highest loadings on the latent variable for the questionnaire with six response categories and the items 5 ("I weigh a lot.") on the latent variable for the questionnaire with two response categories scale. Here, 50.0% of items loaded higher than the gold-standard physical weight report for both scales.

For age, the items 5 ("I've already lived most of my life.") and 7 ("Over time, my memory has deteriorated.") had the highest loadings on the latent variable for the questionnaire with six response categories and the items 10 ("My whole life still lies ahead of me.") and 5 for the questionnaire with two response categories. Overall, 30.0% (six categories) respectively. 40.0% (two categories) of the items loaded higher than the gold-standard physical age report. **TODO**

The items loadings for both weight and age scales suggest that the gold-standard items were not essential for measuring the latent variable.

>>>>

First, we constructed items for the physical characteristics of body height, body weight, and age. Age was defined as chronological age or age in years, corresponding to the time elapsed since a person's birth [@montepareYouReOnly1989; @schwallDefiningAgeUsing2012]. Height was defined as the height of an upright person from the sole of the foot to the top of the head in centimeters, and body weight as the physical mass of a person in kilograms [@martinAnthropometrieAnleitungSelbstaendigen1929]. The items were developed by 124 psychology students [deductive item construction, @burischApproachesPersonalityInventory1984] and 24 persons [prototype approach, see @broughtonPrototypeStrategyConstruction1984] who deviate from the German population mean at least one standard deviation in the relevant characteristics (separately for women and men). The 24 persons were asked to think of prototypical behaviors for each construct to assess the whole spectrum of the latent variable. This process produced a total of 138 items (response categories were not tested). In a second step, we examined how these items were interpreted and whether they were connected to ideas unrelated to the constructs using cognitive interviews. We conducted eight interviews with a length of two to three hours and applied various cognitive procedures, including probing, paraphrasing, concurrent-think-aloud, and retrospective-think-aloud [@prueferVerfahrenEvaluationSurvey1996; @prueferKognitiveInterviews2005]. This should be a sufficient number as recommendations suggest five to 30 interviews and indicate that the most serious problems can already be identified with a small number of interviews [@willisCognitiveInterviewing2005]. The interviews were conducted with `r nrow(dat_interviews)` people from the target group (`r nrow(dat_interviews[dat_interviews$Geschlecht == 1,])` women and `r nrow(dat_interviews[dat_interviews$Geschlecht == 2,])` men) aged between `r min(dat_interviews$Alter)` and `r max(dat_interviews$Alter)` (*M* = `r printnum(mean(dat_interviews$Alter))`, *SD* = `r printnum(sd(dat_interviews$Alter))`). The level of education ranged from high school diploma to university degree. After the cognitive interviews, misleading items were reformulated or eliminated [@faulbaumWasIstGute2009], resulting in 61 remaining items. In a third step, these 61 items were tested online on a new sample, which initially consisted of 456 people. However, we excluded `r 456 - nrow(dat_pilot)` subjects because they had not completed the questionnaire, resulting in a sample of `r nrow(dat_pilot)` participants aged between `r min(dat_pilot$Alter)` and `r max(dat_pilot$Alter)` years (*M* = `r printnum(mean(dat_pilot$Alter))`, *SD* = `r printnum(sd(dat_pilot$Alter))`), including `r nrow(dat_pilot[dat_pilot$Geschlecht == 1,])` women (`r printnum(nrow(dat_pilot[dat_pilot$Geschlecht == 1,]) / nrow(dat_pilot) * 100)` %) and `r nrow(dat_pilot[dat_pilot$Geschlecht == 2,])` men (`r printnum(nrow(dat_pilot[dat_pilot$Geschlecht == 2,]) / nrow(dat_pilot) * 100)` %). The level of education was distributed as follows: `r round(nrow(dat_pilot[dat_pilot$Bildungsniveau == 1,]) / nrow(dat_pilot) * 100, 2)` % no school leaving certificate, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 2,]) / nrow(dat_pilot) * 100)` % secondary school leaving certificate/elementary school or equivalent, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 3,]) / nrow(dat_pilot) * 100)` % secondary school or equivalent, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 4,]) / nrow(dat_pilot) * 100)` % vocational baccalaureate or high school diploma, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 5,]) / nrow(dat_pilot) * 100)` % college degree or university degree, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 6,]) / nrow(dat_pilot) * 100)` % doctorate or habilitation. Out of all respondents, `r printnum(nrow(dat_pilot[dat_pilot$Muttersprache == 1,]) / nrow(dat_pilot) * 100)` % stated German as their native language. Based on these survey responses, we excluded items if a) their main item loading was not on the intended factor, b) their main and secondary loadings were almost equal, or c) they had loading below or equal to .30. Additionally, if (d) items had very similar content, items with lower loadings were excluded. Details on the process of item construction can be found in @eichhornItem2019. We reproduced the exploratory factor analyses and item total correlations and report them in our electronic online materials (TODO). Based on these criteria 12 items were selected for each construct (see @tbl-5). For each of the three questionnaires, two versions were created – one with a 6-point scale with verbal endpoints ("does not apply" to "applies", only endpoints were labeled) and fully numerical anchoring for the dichotomous response format with "applies" or "does not apply".

@@

Both response formats exhibited similar loading patterns, but those with two response categories obtained slightly higher loadings compared to six response categories for most items. On average, questionnaires with two response categories showed slightly better fit indices than those with six categories. All model fits were within the expected range for questionnaire items [@goretzkoEvaluatingModelFit2024]. In particular, the fit indices for the questionnaires on body height and weight with two response categories showed the best fit.

```{r}
#| eval: false
higher_load <- results$load_df_nomod |> group_by(measure, categories) |>
  reframe(freq = sum(head(abs(est.std), 12) > tail(abs(est.std), 1)))
```

For the body height scales, the self-reported physical height loaded higher or similarly high on the latent variables compared to questionnaire items.
However, the pattern was different the weight and age scales, where several items loaded substantially higher on the latent variable than the physical items:
For weight, the items 4 ("I am obese.") and 5 ("I weigh a lot.") had especially high loadings on the latent variable for the questionnaire with with two and six response categories.
For age, item 7 ("My pants have a small waistband.") had especially high loadings on the latent variable for the questionnaire with two and six response categories.

The item loadings for the weight and age scales suggest that the gold-standard items were not essential for measuring the latent variable.
